% !TEX program = pdflatex
% !TEX spellcheck = en_US
\documentclass[aspectratio=169,svgnames,xcolor=table,t]{beamer}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{listings}
\lstset{basicstyle=\ttfamily,escapechar=\%,escapeinside={\#(}{\#)}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath,amssymb,amstext}
\usepackage{mathtools}
\usepackage{calc}
\usepackage{tikz}
\usetikzlibrary{positioning}
\tikzset{place/.style={draw}}
\tikzset{>=stealth, auto, node distance=2.5cm, every loop/.style={->, min distance=10mm, in=0, out=60, looseness=10}}
\usepackage{pgfplots}
\usepackage{algorithm, algorithmic}
\usepackage{caption}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{uno_style.tex}
\input{macro.tex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Use enumitem and restore beamer defaults
% \usepackage{enumitem}
% \setitemize{label=\usebeamerfont*{itemize item}%
%   \usebeamercolor[fg]{itemize item}
%   \usebeamertemplate{itemize item}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[J. Fandinno \& Eleuterio Lillo Portero]{{\usebeamercolor[fg]{alerted text}Jorge Fandinno} and Lute Lillo}

\date{AAAI 2025}

\institute[UNO]{
    Department of Computer Science \\
    University of Nebraska Omaha
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Document Metadata
\title{Solving Epistemic Logic Programs using\\Generate-and-Test with Propagation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\frame{\titlepage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Epistemic Logic Programs}
    \begin{myitemize}
        \item Epistemic logic programs are an extension of Answer Set Programs (ASP) with epistemic operators that allow reasoning about the knowledge of agents. For example,
        \begin{align*}
            \mathit{felon}   &\leftarrow \K\, \mathit{break\_rule}
            \\
            \mathit{suspect} &\leftarrow \neg \K\, \mathit{break\_rule}, \neg \K \neg \mathit{break\_rule}
        \end{align*}
        This encodes that a person is~$\mathit{felon}$ if we can determine that she broke a rule, and she is a~$\mathit{suspect}$ if it cannot be determined neither that she has broken the law nor that she has not done it.
    
        \bigskip

        \item \hi{Deciding} if an epistemic logic program has an answer (called \hi{worldview}) \hi{is one lever higher than ASP}: $\Sigma^P_3$-complete and~$\Sigma^P_2$-complete for the case of disjuntive\nobreak-free programs.
        
        \bigskip

        \item Epistemic logic programs provide a more natural way to model problem in the second level of the polynomial hierarchy than disjunctive ASP with saturation techniques.
    \end{myitemize}
    \end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{In the paper}
    \begin{myitemize}
        \item We formalize an \hi{abstract generate-and-test framework} for computing worldviews that covers existing solvers such as \texttt{EP-ASP} or \texttt{eclingo}:
        \begin{itemize}
            \item Generate-and-test approaches use a generator program to generate candidates and a test program to check if a candidate corresponds to a worldview.

            \item We provide \hi{sufficient conditions} on the generator and test programs for the \hi{correctness} of the algorithm.
        \end{itemize}
        
        \medskip

        \item We propose a new generator program that \hi{propagates the consequences of epistemic decisions to reduce the number of candidates} that need to be tested.
        \begin{itemize}
            \item Each candidate to be tested requires a linear number of calls to an ASP solver.
        \end{itemize}

        \medskip

        \item We implement a new solver based on this new generator
        \begin{itemize}
            \item benchmarks show that it solves \hi{87\% more instances} than existing solvers and is \hi{\speedup{3.3x} faster}.
        \end{itemize}
    \end{myitemize}
    \end{frame}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{frame}{Solving Epistemic Logic Programs}
% \begin{myitemize}
%     \item \hi{\texttt{eclingo}} is a solver for epistemic logic programs.
%     \begin{align*}
%         \mathit{felon}   &\leftarrow \K\, \mathit{break\_rule}
%         \\
%         \mathit{suspect} &\leftarrow \neg \K\, \mathit{break\_rule}, \neg \K \neg \mathit{break\_rule}
%     \end{align*}
%     This encodes that a person a~$\mathit{felon}$ if we can determine that she broke a rule, and she is a~$\mathit{suspect}$ if it cannot be determined that she has broken nor that she has not done.

%     \item Deciding if an epistemic logic program has an answer (called \hi{worldview}) is~$\Sigma^P_3$-complete and~$\Sigma^P_2$-complete for the case of disjuntive\nobreak-free programs.
    
%     \item We developed a new version of \texttt{eclingo} that is \speedup{3.3x} faster and solves 87\% more instances.
% \end{myitemize}
% \end{frame}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{frame}{Solving Epistemic Logic Programs}
%     \begin{myitemize}
%         \item[] As the previous version, \texttt{eclingo} 2.0 uses a generate-and-test approach.
%     \end{myitemize}
%     \begin{algorithm}[H]
%         \scriptsize
%         \caption{Generate-and-test computation of~$n$ worldviews of a program~$\Pi$ in normal form.}\label{alg:guess-and-check}
%         \hspace*{\algorithmicindent} \textbf{Input} Generate program $G(\Pi)$\\
%         \hspace*{\algorithmicindent} \textbf{Input} Test program $T(\Pi)$\\
%         \hspace*{\algorithmicindent} \textbf{Input} Number of requested worldviews~$n$\\
%         \hspace*{\algorithmicindent} \textbf{Output} Set~$\Omega$ containing~$n$  worldviews of~$\Pi$
    
%         \begin{algorithmic}[1] %[1] enables line numbers
%             \STATE Let $\Omega=\emptyset$.
%             \FOR{$M$ in $\SM{G(\Pi)}$} \label{alg.generator.loop.ini}
%                 \IF {\texttt{Test$(T(\Pi),\,M)$}}
%                 \STATE $\wv$ = \texttt{BuildWorldView$(M)$}
%                 \STATE $\Omega$ = $\Omega \cup \{ \wv \}$
%                     \IF {$|\Omega| \geq n$}
%                     \STATE \textbf{return} $\Omega$
%                     \ENDIF
%                 \ENDIF
%             \ENDFOR \label{alg.generator.loop.end}
%             \STATE \textbf{return} $\Omega$
%         \end{algorithmic}
%     \end{algorithm}
% \end{frame}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Normal Form and Basic Generator Program}
    \begin{myitemize}
        \item \hi{Normal form}: We first compute a normal form of the program in which negation is not in the scope of the epistemic operator using auxiliary atoms.
        
        \item \hi{Basic generator program}: Replaces epistemic atoms~$\K\,a$ by a fresth atom~$\kk{a}$ and introduces a choice rule~$\{ \kk{a} \}$ for each fresh atom.
    \end{myitemize}
    For epistemic program
    \begin{align*}
        a && 
        b &\leftarrow \K a
        &
        c &\leftarrow \K b
    \end{align*}
    we get the generator program
    \begin{align*}
        a 
        && 
        b &\leftarrow ka
        &
        c &\leftarrow kb
        &
        \{ \kk{a} \} &\leftarrow a
        &
        \{ \kk{b} \} &\leftarrow b
    \end{align*}
    \pause
    The original program has unique worldview~$[\{a,b,c\}]$, while the generator program produces three candidates: $\{a\}$, $\{a,\kk{a},b\}$ and~$\{a,\kk{a},b,\kk{b},c\}$. Only the last candidate corresponds to its unique worldview.
\end{frame}

\newcommand{\lenitem}[2][.5\linewidth]{\parbox[t]{#1}{\strut #2\strut}}


% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Generate-and-Test Framework}
    \mbox{}\hfill\raisebox{20pt}[0pt][0pt]{%
        \begin{minipage}[t]{.4\linewidth}
            \begin{algorithm}[H]
                \footnotesize
                \captionsetup{font=small}
                \caption{Generate-and-test.}
                \hspace*{\algorithmicindent} \textbf{Input} Generate program $G(\Pi)$\\
                \hspace*{\algorithmicindent} \textbf{Input} Test program $T(\Pi)$
                % \\
                % \hspace*{\algorithmicindent} \textbf{Input} Number of requested worldviews~$n$
                % \\
                % \hspace*{\algorithmicindent} \textbf{Output} Set~$\Omega$ containing~$n$  worldviews of~$\Pi$
            
                \begin{algorithmic}[1] %[1] enables line numbers
                    \FOR{$M$ in $\SM{G(\Pi)}$}
                        \IF { \text{$\SMC{T(\Pi); k(M)}$ matches $M$} }
                        \STATE \textbf{return} $\SM{T(\Pi)}$
                        \ENDIF
                    \ENDFOR
                    \STATE \textbf{return} \text{No worldview}
                \end{algorithmic}
            \end{algorithm}%
            \vspace{-10pt}
            \scriptsize%
            $\SMC{T(\Pi); k(M)}$ denotes the cautions consequences of~$T(\Pi)$ under the assumptions of the epistemic atoms in~$M$.
        \end{minipage}%
    }%
    \vspace*{-\baselineskip}
    \begin{myitemize}
        \item \lenitem{A \hi{generator program}}
        \begin{itemize}
            \item \lenitem{has an answer set for each world view of the epistemic program where~$\kk{a}$ is true if~$\K a$ is true in the world view.}
            \item  \lenitem{Each of its answer set corresponds to a candidate that is not empty.}
        \end{itemize} 
        \item \lenitem{Fixing the epistemic atoms of the \hi{tester program} produces as stable models all the elements of the worldview.}
    \end{myitemize}

    \vspace{10pt}

    \begin{theorem}
        The generate-and-test algorithm produces all worldviews of a program~$\Pi$ if the generator and test programs satisfy the above conditions.
    \end{theorem}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Propagate Epistemic Consequences}
    \vspace{-20pt}
    \begin{align*}
        a &&
        b &\leftarrow \K a
        &
        c &\leftarrow \K b
    \end{align*}
    To see that~$[\{a,b,c\}]$ is the unique worldview of this program, we can reason as follows:
    \begin{itemize}
        \item $\K a$ is true because~$a$ is a fact, so it must be true in all stable models.
        \item From~$\K a$ we can derive~$b$ from the second rule. 
        \only<1>{This depends on the world view and not on the particular stable model in it. So~$b$ must be true in all stable models and, thus,~$\K b$ follows.}
        \item From~$\K b$ we can derive~$c$ in a similar way.
    \end{itemize}
    
    \pause
    \pause
    \begin{myitemize}
        \item The key idea is to propagate the consequences of epistemic decisions.
        \item We can add rules that reflect this reasoning inside the generator program.
    \end{myitemize}
    \pause
    \begin{align*}
        a 
        && 
        b &\leftarrow ka
        &
        c &\leftarrow kb
        &
        \{ ka \} &\leftarrow a
        &
        \{ kb \} &\leftarrow b
        \\
        \hi{\kp{a}}&
        &
        \hi{\kp{b}} &\hi{\leftarrow \kk{a}}
        &
        \hi{\kp{c}} &\hi{\leftarrow \kk{b}}
        &
        \mathrlap{\hspace{-16pt}\hi{\leftarrow \kp{a} \wedge \neg ka}}
        &&
        \mathrlap{\hspace{-16pt}\hi{\leftarrow \kp{b} \wedge \neg \kk{b}}}
    \end{align*}
    \hi{$\kp{b}$} is used to distinguish that~$\K b$ is a consequence derived from choices of epistemic atoms while~$kb$ only states that~$\K b$ has been chosen to be true.
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Propagate Epistemic Consequences}
    \begin{myitemize}
        \item[] For every normal rule of the form
        \begin{align}
            a_1 &\leftarrow 
            a_2,\!\dotsc\!,\,a_k,
            \neg a_{k+1},\!\dotsc\!,\,\neg a_m,\,
            {L_{m+1}},\!\dotsc\!,\,{L_n}
        \end{align}
        with each~$L_i$ a~$\K$-atom, we add a rule of the form
        \begin{gather}
            \begin{aligned}
                \kp{a_1} &\span\span\leftarrow 
                \kp{a_2},\dotsc,\,\kp{a_k},\,
                \kpnott{a_{k+1}},\dotsc,\,\kpnott{a_m},\,
                L_{m+1},\dotsc,\,L_n
            \end{aligned}
        \end{gather}
        \vspace*{-15pt}
        \begin{itemize}
            \item \kp{a} means that we have concluded that~$a$ is true in all stable models corresponding to the considered candidate worldview.
            \item \kpnott{a} means that we have concluded that~$a$ is false in all stable models corresponding to the considered candidate worldview.
            \item atoms of the form~\kpnott{a} are derived through a kind of completion of the program.
        \end{itemize}
    \end{myitemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Propagate Epistemic Consequences}
    For program
    \begin{align*}
        b &\leftarrow \neg a
        &
        c &\leftarrow \K b
    \end{align*}
    we obtain the generator program
    \begin{align*}
        &&
        b &\leftarrow \neg a
        &
        c &\leftarrow kb
        &
        \{ \kk{b} \} &\leftarrow b
        \\
        \hi{\kpnott{a}}&
        &
        \kp{b} &\leftarrow \hi{\kpnott{a}}
        &
        \kp{c} &\leftarrow \kk{b}
        &&
        \mathrlap{\hspace{-16pt}\leftarrow \kp{b} \wedge \neg \kk{b}}
    \end{align*}
    which produces a single candidate~$\{\kp{a},b,\kp{b},\kk{b},c,\kp{c}\}$ that corresponds to the unique worldview~$[\{b,c\}]$.

    \begin{itemize}
        \item $\kpnott{a}$ is a fact because there is no rule with~$a$ in the head.
        \item The translation uses auxiliary atoms for each body to avoid the quadratic blowup of the completion.
    \end{itemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[c]{Propagate Epistemic Consequences}
    \begin{theorem}
        This method produces a valid generator program, and, thus, it can be used in our generate-and-test framework to correctly compute the worldviews of a program.
        \\
        Furthermore,
        \begin{itemize}
            \item It produces at most as many candidates as the basic generator program.
            \item For some program classes, it produces exponentially fewer candidates.
            \item It only introduces a linear overhead.
        \end{itemize}
    \end{theorem}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[c]{Implementation}
    \begin{myitemize}
        \item[] Our implementation is built on top of version~5.7 of the ASP solver~\texttt{clingo} using Python and ASP.%
        \begin{myitemize}
            \item It supports most language features of \texttt{clingo} such as choice rules, aggregates, intervals, etc.
            
            \item The code is available with MIT license at \hi{\url{https://github.com/potassco/eclingo}}.
            
            \item It replaces the previous version of \hi{eclingo}.
                        
            \item We use reification and \hi{metaprogramming} to describe the generate and test programs.
            
            \item This allow developing new solvers with different generate and test programs by \hi{only modifying the ASP metaprogram}.
        \end{myitemize}
    \end{myitemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Experimental Results and Survival Plot}
    \thispagestyle{empty} 
    \begin{columns}[T]
        \begin{column}{0.45\textwidth}
            \begin{myitemize}
                \small
                \item It solves 87\% more instances than existing solvers and is \speedup{3.3x} faster.

                \bigskip

                \item It dominates all the other solvers in all benchmarks.
                
                \bigskip
                
                \item \texttt{EP-ASP} has a specific mode for conformant planning that uses domain-specific heuristics.
                \begin{itemize}
                    \item Our solver solves \speedup{50\%} more instances despite not using domain-specific heuristics.
                \end{itemize}
            \end{myitemize}
        \end{column}
        \begin{column}{0.5\textwidth}
            \hspace*{-15pt}
            \input{solver_comp.tex}
        \end{column}
    \end{columns}
    % \begin{myitemize}
    %     \item We compare our new solver with~\texttt{EP-ASP},
    %     the previous version of~\texttt{eclingo},
    %     \texttt{selp}, and~\texttt{elp2qasp}.
    % \end{myitemize}
    % \vspace*{-20pt}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{frame}{Experimental Results (Highlights)}
%     \begin{myitemize}
%         \item Our solver is \speedup{3.3x} faster than the previous version and it solves 87\% more instances.
%         \item It dominates the other solvers in all benchmarks.
%         \begin{itemize}
%             \item In the eligible benchmark, it solves 100\% of the instances and it is \speedup{7.5x} faster than the previous version.
            
%             \item In the Yale benchmark, it solves every instance in less than 1 second (same as the previous version).
%             \begin{itemize}
%                 \item All but \texttt{elp2qasp} solve all instances.
%             \end{itemize}
            
%             \item In the bomb benchmark, it is \speedup{2.4x} faster than~\texttt{EP-ASP} and it solves almost 3 times more instances.
%             \begin{itemize}
%                 \item Compared to the previous version, it is \speedup{2.9x} faster and it solves almost 7.7 times more instances.
%                 \item \texttt{EP-ASP} has a specific mode for conformant planning that uses domain-specific heuristics. When compared to this version, our solver is \speedup{1.5x} faster and solves \speedup{50\%} more instances.
%             \end{itemize}
%         \end{itemize}
%     \end{myitemize}
% \end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}{Conclusions}
    \begin{myitemize}
        \item We have developed a theoretical framework for generate-and-test computation of worldviews of epistemic logic programs and prove its correctness.
        
        \item We have instantiated this framework with a new generator program that propagates the consequences of epistemic decisions.
        \begin{itemize}
            \item We prove that can exponentially reduce the number of candidates while only introducing a linear overhead. 
        \end{itemize}

        \item We implemented a new solver which solves 87\% more instances than state-of-the-art solvers and \speedup{3.3x} faster.
        
        \bigskip
        
        \item \hi{Future work}: we plan to study how to introduce ideas from conflict-driven learning to further speed up solving epistemic logic programs.
    \end{myitemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\frame{\backcoverpage}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}